Wait-free Queues with Polygarithmic Step Complexity

### 2022-08-17 article/psuedocode.tex
review indexDeq proof

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList}}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\end{itemize}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
\If{\tt{dir {\keywordfont is} right}}
%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerRight}
\EndIf
\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
\EndIf
\EndFunction{IndexDeq}

@@

\begin{itemize}
\item Prefix of the $op$ in the sequence of operations $S$, is the operations strictly before $op$.
  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])} ordered by their process id.
  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])} ordered by their process id.
  \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
\item Order of the enqueue operations in the node $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
\item Order of the dequeue operations in the node $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
\end{itemize}
\end{definition}
\it{Note that in the non-root nodes we only need ordering of enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}

@@

\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/doubling.png}

  \caption{Distance relations between $b,c,end$}
  \label{fig::doubling}
\end{figure}

@@

@@

\begin{lemma}[Index correctness]
 \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case \nf{root.IndexDEQ(b,i)} is trivial (Line \ref{indexBaseCase}).
  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}). After that the order in $D(n.parent, superblock)$ is computed and \nf{index()} is called on \nf{n.parent} recursively. Then if the operation was propagated from the right child the number of dequeues from the left child are added to it (Line \ref{considerRight}), because the left child operations come before the right child operations (Definition \ref{ordering}).
\end{proof}
\textit{Do I need to talk about the computation of the order in the parent which is based on the definition of ordering of dequeues in a block? Make sure to show preconditions of all invocation of \nf{BSearch} are satisfied.}

@@

\begin{lemma}[Computing SuperBlock]\label{superBlock}
After computing line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)}, \nf{n.parent.blocks[superblock]} contains $D(n,b,i)$.
\end{lemma}
\begin{proof}
Lemmas 28,29,30,31.
\end{proof}
\begin{lemma}
Value read for \nf{super[b.group]} in line 418 is not null.\end{lemma}
\begin{proof}
  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}. So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading.
\end{proof}
\begin{lemma} \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
\end{lemma}
\begin{proof} Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.\end{proof}

@@

% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
\begin{lemma} \nf{super[i+1]-super[i]}$\leq p$\end{lemma}
\begin{proof}
 In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49. Assume the current value of the counter in node \nf{n} is \nf{i+1} and still \nf{super[i+1]} is not set. If an instance of successful \nf{Refresh(n)} finishes \nf{super[i+1]} is set a new value and a block is added after \nf{n.parent[sup[i]]}. There could be at most $p$ successful unfinished concurrent instances of \nf{Refresh()} that have not reached line 49. So the distance between \nf{super[i+1]} and \nf{super[i]} is less than $p$.
 \begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/index2}
\end{figure}  
\end{proof}
\begin{lemma}[super property]\label{superCounter}
If \nf{super[i] $\neq$ null} in node \nf{n}, then \nf{super[i]} is the index of the superblock of a block with \nf{time=i} in \nf{n.parent.blocks}.
\end{lemma}
\begin{lemma}\label{superRange} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
\end{lemma}

@@

>>>>

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList}}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
\end{itemize}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}

\If{\tt{dir {\keywordfont is} left}}
\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-left}} \cmt{consider the enqueues in the previous blocks from the left child} \label{considerPreviousLeft}
\EndIf

\If{\tt{dir {\keywordfont is} right}}
\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-right}} \cmt{consider the enqueues in the previous blocks from the right child} \label{considerPreviousRight}
\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerLeftBeforeRight}
\EndIf
\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
\EndIf
\EndFunction{IndexDeq}

@@

\begin{itemize}
\item Prefix of the $op$ in the sequence of operations $S$, is the operations strictly before $op$.
  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])}. $E(leaf,b)$ is the enqueue operation in \nf{ops(leaf.blocks[b])}.
\begin{align*} 
E(n,b) =&  E(n.left,n.blocks[b-1].end\sub{left}+1).E(n.left,n.blocks[b-1].end\sub{left}+2)...E(n.left,n.blocks[b].end\sub{left}). \\ 
&E(n.right,n.blocks[b-1].end\sub{right}+1).E(n.right,n.blocks[b-1].end\sub{right}+2)...E(n.right,n.blocks[b].end\sub{right})
\end{align*}
  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
\item Order of the enqueue operations in the node $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])}. $D(leaf,b)$ is the dequeue operation in \nf{ops(leaf.blocks[b])}.
\begin{align*} 
D(n,b) =&  D(n.left,n.blocks[b-1].end\sub{left}+1).D(n.left,n.blocks[b-1].end\sub{left}+2)...D(n.left,n.blocks[b].end\sub{left}). \\ 
&D(n.right,n.blocks[b-1].end\sub{right}+1).D(n.right,n.blocks[b-1].end\sub{right}+2)...D(n.right,n.blocks[b].end\sub{right})
\end{align*}
    \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
\item Order of the dequeue operations in the node $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
\end{itemize}
\end{definition}
\it{Note that in the non-root nodes we only need ordering of enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}

@@

\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/doubling.png}
  \caption{Distance relations between $b,c,end$}
  \label{fig::doubling}
\end{figure}

@@

Hello Im hoseen master student in cs at York uni. I will defend my thesis by Nov this year and I was looking for ful time jobs starting from novomber. I have experince in both acdemia and indusrty. When I was a bachelors srudent I workd from 2 years as a Software dveloper. I have expereince with backend devlopment. I have implemented, deploeyd an maintained backend systems.

@@

@@

\begin{lemma} Let $n.propagates$ be the number of groups of blocks that have been propagated from node \nf{n} to its parent (successful \nf{n.parent.Refresh()}).
We have \nf{num\sub{propagated}}$\leq n.propagates\leq$\nf{num\sub{propagated}+p}. \nf{p} is the number of processes.
\end{lemma}
\begin{proof}
    \tt{num\sub{propagated}} is incremented after propagating (Line \ref{incNP}). Since maybe some process falls sleep before incrementing \tt{num\sub{propagated}} it may be behind by \nf{p}.
\end{proof}

\begin{lemma} \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
\end{lemma}
\begin{proof} Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.\end{proof}

@@

% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
\begin{lemma} \nf{super[i+1]-super[i]}$\leq p$\end{lemma}
\begin{proof}
 In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49. Assume the current value of the counter in node \nf{n} is \nf{i+1} and still \nf{super[i+1]} is not set. If an instance of successful \nf{Refresh(n)} finishes \nf{super[i+1]} is set a new value and a block is added after \nf{n.parent[sup[i]]}. There could be at most $p$ successful unfinished concurrent instances of \nf{Refresh()} that have not reached line 49. So the distance between \nf{super[i+1]} and \nf{super[i]} is less than $p$.
 \begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/index2}
\end{figure}  
\end{proof}
\begin{lemma}[super property]\label{superCounter}
If \nf{super[i] $\neq$ null} in node \nf{n}, then \nf{super[i]} is the index of the superblock of a block with \nf{time=i} in \nf{n.parent.blocks}.
\end{lemma}
\begin{lemma}\label{superRange} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.group]}.
\end{lemma}

@@

\begin{lemma}[Computing SuperBlock]\label{superBlock}
For the \nf{superblock} value computed in line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)} we have \nf{n.parent.blocks[superblock]} contains $D_{n,b,i}$.
\end{lemma}
\begin{proof}
First we show the value read for \nf{super[b.group]} in line 418 is not null.
  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}. So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading. Then by Lemma \ref{superRange}if we search in the range $p$, we can find the superblock.

\end{proof}

\begin{lemma}[Index correctness]
 If \nf{n.blocks[b].num\sub{enq}}$\geq$\nf{i} then \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case where \nf{n} is root is trivial (Line \ref{indexBaseCase}).
  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}). After that the order in $D(n.parent, superblock)$ is computed. Note that by Lemma \ref{blockSize} in each block there is at most one operation from each process and operations of one type are ordered based on the order in the subblocks (See Figure \ref{fig::orderFromSubblocks}). Finally \nf{index()} is called on \nf{n.parent} recursively and it returns the correct response from induction hypothesis. If the operation was propagated from the right child the number of dequeues from the left child are added to it (Line \ref{considerRight}), because the left child operations come before the right child operations (Definition \ref{ordering}).
\end{proof}
\textit{Make sure to show preconditions of all invocation of \nf{BSearch} are satisfied.}

\begin{figure}[hbt]  
  \center\includegraphics[width=4in]{pics/orderFromSubblocks.png}
  \caption{Relation of ordering of operations of a block from its subblocks}
  \label{fig::orderFromSubblocks}
\end{figure}
